Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Edits in sections 2.1 and 2.2

This section presents the key aspects of the design of our type system,
called \ourTypeSystem, in the context of a core calculus for an
object-oriented language.

@@

\Cref{fig:grammar} formalizes these notions.
\Cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
lacks a top-level type, but any type arguments are full types
$\tau$.  This paper does not formally define basetype, and uses $\beta$,
for simplicity of exposition.
In our core calculus, a type $\tau$ always has a qualifier.
Our implementation, \theDeterminismChecker, uses inference and defaults to
permit users to omit type qualifiers\todo{Forward reference}.

A type checker verifies the types written in a program.
A type qualifier constrain the set of possible run-time values, that is,
$\kappa \ /beta \sqsubseteq \beta$.
As a result, a qualifier type system does not allow any behavior that the
original type system does not.
However, the qualifier type system may reject more programs, and thus
affords stronger guarantees.
(Our implementation \theDeterminismChecker runs the underlying base type
system, then issues additional errors and warnings.)
\todo{Some qualifier type systems can be defined independently of the underlying
type system.  We don't do that because there are interactions between the
basetypes and the type qualifiers.  This allows greater precision, which
is important in practice.}

@@

    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
    \quad\quad
    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$

    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
    type qualifiers, types, and unqualified types $\beta$.}
    \label{fig:typecheck-rules}
\end{figure}

    $\infer[\rulename{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

@@

\OurTypeSystem checks all the standard typing rules including the ones shown in \cref{typecheck-rules-standard} with
the only difference of having an additional type qualifier on each of the basetypes.

@@

The core of the determinism type system is the following type qualifiers:
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that the expression evaluates to equal values in
      all executions; for an unordered collection such as a set, iteration
      also yields the values in the same order.
\end{itemize}
\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.

\<OrderNonDet> may only be written on collections and maps.

A map is a dictionary or associative array, such as a hash table.
\OurTypeSystem largely treats a map as a collection of key--value pairs.

Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including qualifiers.

@@

    $\infer[\rulename{ordernondet collection}]
    {\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}
    {\kappa_c \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_c &
      \beta_c \sqsubseteq \<Collection> & \wellformed{\kappa_e \ \beta_e}}$

@@

    $\infer[\rulename{nondet collection}]
    {\wellformed{\|NonDet| \ C\angles{\|NonDet| \ \beta_e}}}
    {C \sqsubseteq \<Collection> & \wellformed{\|NonDet| \ \beta_e}}$

@@

%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_a }$
%     
%     \bigskip
%     
%     $\infer[\rulename{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
%     
%     \bigskip

    $\infer[\rulename{ordernondet map}]
    {\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}
    {\kappa_m \sqsubseteq \|OrderNonDet|
      & \kappa_k \sqsubseteq \kappa_m
      & \kappa_v \sqsubseteq \kappa_m
      & \beta_m \sqsubseteq \<Map>
      & \wellformed{\kappa_k \ \beta_k}
      & \wellformed{\kappa_v \ \beta_v}
    }$

@@

    $\infer[\rulename{nondet map}]
    {\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}
    {\beta_m \sqsubseteq \<Map> & \wellformed{\|NonDet| \ \beta_k} & \wellformed{\|NonDet| \ \beta_v}}$

    \caption{Type well formedness}

%\todo{\todo{I think this will be OK, by redefining $\beta$.}
%  The collection rules do not permit nested collections; similarly, the
%  (currently commented out) array rules did not permit any two-dimensional
%  array to be well-formed.  There is a bigger issue here.
%  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
%  to use a different symbol for it, probably $c$), but $\beta_e$ should be
%  a full type, which might contain type qualifiers within it.  (Actually,
%  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
%  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
%  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
%  is well-formed.  Likewise, antecedents are needed for the map rules.
%  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
%  where $\tau'$ is itself not well-defined.}

@@

\begin{figure}
    \centering
    \begin{tabular}{|l|l|l|l|l|}
        \cline{3-5}
        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Element type} \\ \cline{3-5}
        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
                        & NonDet      &  valid     &  invalid    & invalid  \\ \cline{2-5}
        Collection type & OrderNonDet &  invalid   &  valid      & valid    \\ \cline{2-5}
                        & Det         &  invalid   &  invalid    & valid    \\ \hline
    \end{tabular}
    \caption{Valid Collection declarations.  The Collection's type qualifier
        must be a supertype or equal to the element type.}
    \label{fig:determinism-collections}
\end{figure}

@@

    \rulename{iteration}: 
    $
    \forallt{\tau} \|OrderNonDet \ C| \angles{\tau} \rightarrow 
    \|OrderNonDet \ Iterator| \angles{\tau}
    $

@@

    \rulename{next element}:
    $
    \forallt{kappa \beta} \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
    \|NonDet|\ \beta
    $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
    \|Det \ boolean|
    $
    \bigskip

@@

    \rulename{size}:
    $
    \forallt{\tau} \|OrderNonDet \ C|\angles{\tau} \rightarrow
    \|Det \ int|
    $

@@

\begin{figure}
    $\infer[\rulename{ordernondet polyup}]
    {\up{\tau} = \|NonDet \ \beta|}
    {\tau = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polyup}]
    {\up{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \|Det \ \beta|}
    {\tau = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
\todo{I think this needs to do something to nested collections.  Or maybe
  just clarify that if there is a nested collection such that this would
  yield an invalid type, then the instantiation does not apply.  That might
  be cleanest.}
\label{fig:poly-resolutions}
\end{figure}

@@

\todo{We don't need the OrderNonDet instantiation.  The other two
  instantiations are enough.
  We don't need the following instantiation:}
\begin{Verbatim}
   // instantiation: OrderNonDet List<Det String> x _ Int -> NonDet String 
\end{Verbatim}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y) : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \|OrderNonDet \ List| & \kappa_{y} \ \beta_{y} \nsqsubseteq \|OrderNonDet \ List| }$

@@

    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y) : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta|}$

>>>>

This section presents the key aspects of our type system,
called \ourTypeSystem, in the context of a core calculus for an
object-oriented language.

@@

\Cref{fig:grammar} formalizes these notions.
For simplicity of exposition,
\cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
lacks a top-level type, but any type arguments are full types
$\tau$.
In our core calculus, a type $\tau$ always has a qualifier.
Our implementation, \theDeterminismChecker, uses inference (\cref{sec:dataflow-java}) and
defaults (\cref{defaulting}) to
permit users to omit type qualifiers.

A type checker verifies the types written in a program.
\OurTypeSystem checks all the standard typing rules including the ones
shown in \cref{typecheck-rules-standard}.  For \ourTypeSystem, $\tau$
represents a qualified type rather than a basetype.

A type qualifier constrains the set of possible run-time values, that is,
$\kappa \ \beta \sqsubseteq \beta$.
As a result, a qualifier type system does not allow any values that the
original type system does not.
However, the qualifier type system may reject more programs, and thus
affords stronger guarantees.
% (Our implementation \theDeterminismChecker runs the underlying base type
% system, then issues additional errors and warnings.)

Type qualifier systems are sometimes defined independently of the
underlying type system.  Because there are interactions between the
basetypes and the type qualifiers, defining them together improves
precision, which is important in practice.

    $\infer[\rulename{call}]{\Gamma \vdash f(\overline{a_i}) : \tau_r}
    {\Gamma \vdash f : \overline{\tau_i} \rightarrow \tau_r & \Gamma \vdash \overline{a_i : \tau_i}}$

@@

    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
    \quad\quad
    \infer{\kappa\ \beta_1 \sqsubseteq \kappa\ \beta_2}{\beta_1 \sqsubseteq \beta_2}$

    \caption{Subtyping rules for \ourTypeSystem's type qualifiers.  $\sqsubseteq$ is overloaded for classes,
    type qualifiers, types, and unqualified types $\beta$.}
    \label{fig:typecheck-rules}
\end{figure}

@@

@@

The core of the determinism type system is the following type qualifiers:
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that the expression evaluates to equal values in
      all executions; for a collection, iteration
      also yields the values in the same order.
\end{itemize}
\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show the subtyping
relationship among the qualifiers.

\<OrderNonDet> may only be written on collections and maps.
A map is a dictionary or associative array, such as a hash table.
\OurTypeSystem largely treats a map as a collection of key--value pairs.
Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including qualifiers.

@@

    $\infer[\rulename{ordernondet collection}]
    {\wellformed{\kappa_c \  C \angles{\kappa_e \ \beta_e}}}
    {\kappa_c \sqsubseteq \<OrderNonDet> &
      C \sqsubseteq \<Collection> &
      \kappa_e \sqsubseteq \kappa_c &
      \wellformed{\kappa_e \ \beta_e}}$

@@

    $\infer[\rulename{nondet collection}]
    {\wellformed{\<NonDet> \ C\angles{\<NonDet> \ \beta_e}}}
    {C \sqsubseteq \<Collection> & \wellformed{\<NonDet> \ \beta_e}}$

@@

%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a \sqsubseteq \<OrderNonDet> & \kappa_e \sqsubseteq \kappa_a }$
%     
%     \bigskip
%     
%     $\infer[\rulename{nondet array}]{\wellformed{\<NonDet> \ \beta_e \ \  \<NonDet> []}}{}$
%     
%     \bigskip

    $\infer[\rulename{ordernondet map}]
    {\wellformed{\kappa_m \  C \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}
    {\kappa_m \sqsubseteq \<OrderNonDet>
      & C \sqsubseteq \<Map>
      & \kappa_k \sqsubseteq \kappa_m
      & \kappa_v \sqsubseteq \kappa_m
      & \wellformed{\kappa_k \ \beta_k}
      & \wellformed{\kappa_v \ \beta_v}
    }$

@@

    $\infer[\rulename{nondet map}]
    {\wellformed{\<NonDet> \ C\angles{\<NonDet> \ \beta_k, \<NonDet> \ \beta_v}}}
    {C \sqsubseteq \<Map> & \wellformed{\<NonDet> \ \beta_k} & \wellformed{\<NonDet> \ \beta_v}}$

    \caption{Type well formedness}
\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}

%\todo{\todo{I think this will be OK, by redefining $\beta$.}
%  The collection rules do not permit nested collections; similarly, the
%  (currently commented out) array rules did not permit any two-dimensional
%  array to be well-formed.  There is a bigger issue here.
%  Consider the ``nondet collection'' rule.  $C_c$ is a class (we need
%  to use a different symbol for it, probably $c$), but $\beta_e$ should be
%  a full type, which might contain type qualifiers within it.  (Actually,
%  $\<NonDet>\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
%  be an antecedent stating that $\<NonDet> \beta_e$ is well-formed, and
%  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
%  is well-formed.  Likewise, antecedents are needed for the map rules.
%  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
%  where $\tau'$ is itself not well-defined.}

@@

\begin{figure}
    \centering
    \begin{tabular}{|l|l|l|l|l|}
        \cline{3-5}
        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Element type} \\ \cline{3-5}
        \multicolumn{2}{c|}{~}  & \<NonDet>     & \<OrderNonDet> & \<Det> \\ \hline
                        & \<NonDet>      &  valid     &  invalid    & invalid  \\ \cline{2-5}
        Collection type & \<OrderNonDet> &  invalid   &  valid      & valid    \\ \cline{2-5}
                        & \<Det>         &  invalid   &  invalid    & valid    \\ \hline
    \end{tabular}
    \caption{Valid Collection declarations.  The Collection's type qualifier
        must be a supertype or equal to the element type.}
    \label{fig:determinism-collections}
\end{figure}

@@

    \rulename{iteration}: 
    $
    \forallt{\tau} \<OrderNonDet> \ C \angles{\tau} \rightarrow 
    \<OrderNonDet> \ \<Iterator> \angles{\tau}
    $

@@

    \rulename{next element}:
    $
    \forallt{kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
    \<NonDet>\ \beta
    $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
    \<Det> \ \<boolean>
    $
    \bigskip

@@

    \rulename{size}:
    $
    \forallt{\tau} \<OrderNonDet>\ C\angles{\tau} \rightarrow
    \<Det> \ \<int>
    $

@@

\begin{figure}
    $\infer[\rulename{ordernondet polyup}]
    {\up{\tau} = \<NonDet> \ \beta}
    {\tau = \<OrderNonDet> \ \beta
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polyup}]
    {\up{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\<Det>, \<NonDet>\}
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \<Det> \ \beta}
    {\tau = \<OrderNonDet> \ \beta}
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\<Det>, \<NonDet>\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
\todo{I think this needs to do something to nested collections.  Or maybe
  just clarify that if there is a nested collection such that this would
  yield an invalid type, then the instantiation does not apply.  That might
  be cleanest.}
\label{fig:poly-resolutions}
\end{figure}

@@

\todo{We don't need the \<OrderNonDet> instantiation.  The other two
  instantiations are enough.
  We don't need the following instantiation:}
\begin{Verbatim}
   // instantiation: OrderNonDet List<Det String> x _ Int -> NonDet String 
\end{Verbatim}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

@@

    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det>}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$

### 2019-04-04 java-design.tex
Fix LaTeX error
Edits in sections 2.1 and 2.2

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could be instantiated
%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> gets instantiated to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

\begin{figure}
    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

@@

    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{\<Map> and \<Set> creation rules}
    \todo{Are these type well-formedness rules?}
\todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
  deterministic set?}
    \label{fig-creation-rules}
\end{figure}

@@

The return type of \<System.getenv}, which reads an operating system
environment variable, is always \<NonDet>.
The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
the result is \<Det> if the argument is \codeid{"line.separator"}; although
it varies by operating system, many programs including \<diff> hide those
differences from the user.
The result is also deterministic for \codeid{"path.separator"} and
\codeid{"file.separator"}, because these lead to the same behavior given
corresponding environments.
The user can specify input properties that must be passed on the \<java>
command line and thus act like inputs to the program; these are also
treated as deterministic.
The inputs to a program are also treated as \<Det>.

>>>>

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could be instantiated
%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det} or \codeid{@NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> gets instantiated to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

\begin{figure}
    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{\<Map> and \<Set> creation rules}
    \todo{Are these type well-formedness rules?}
\todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
  deterministic set?}
    \label{fig-creation-rules}
\end{figure}

@@

The return type of \<System.getenv>, which reads an operating system
environment variable, is always \<NonDet>.
The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
the result is \<Det> if the argument is \codeid{"line.separator"}; although
it varies by operating system, many programs including \<diff> hide those
differences from the user.
The result is also deterministic for \codeid{"path.separator"} and
\codeid{"file.separator"}, because these lead to the same behavior given
corresponding environments.
The user can specify input properties that must be passed on the \<java>
command line and thus act like inputs to the program; these are also
treated as deterministic.
The inputs to a program are also treated as \<Det>.
